---
title: "Australian Domestic Flights"
author: "Stephen J Parton"
date: "2022-10-20"
categories: [code, analytics, flights, maps]
website:
  sidebar:
    style: "docked"
    search: true
    contents:
      - section: "Data Exploration"
      - index.qmd
format: 
  html: 
    theme: litera
    toc: true
    toc-title: Contents
    number-sections: true
    number-depth: 3
    code-fold: true
    code-summary: "Code"
    code-tools: true
execute: 
  echo: true
  warning: false
  error: false
  freeze: true
  cache: true
---

## Introduction

This analysis summarises Australian domestic flight volumes and on-time performance (OTP) issues by airline over time.

It has been prepared mainly to get more used to Quarto, adn comprises:

-   Initial data load and preprocessing - not shown

-   Domestic Flight Analysis - This document

-   Global Flight Analysis - Next document

-   Forecasting - to come

```{r, load_packages}
#| echo: false

library(tidyverse)


library(scales)

library(leaflet)
library(leaflet.minicharts)


library(janitor)
library(plotly)
library(gganimate)

library(dygraphs)


options(scipen = 999)



```

## Data Sources

Data is sourced from https://data.gov.au/ site, specific datasets used being:

-   Top routes

-   Industry Totals

-   On-Time-Performance - Domestic

(need to add notes/refs)

```{r, data_load}
#| echo: false


top_routes_prep_df <- read_rds("./artifacts/top_routes_prep_df.rds")
ind_totals_prep_df <- read_rds("./artifacts/ind_totals_prep_df.rds")
dom_cargo_prep_df  <- read_rds("./artifacts/dom_cargo_prep_df.rds")
otp_prep_df        <- read_rds("./artifacts/otp_prep_df.rds")

latest_date        <- max(top_routes_prep_df$date)


otp_prep_df <- otp_prep_df %>% 
  mutate(across(airline,str_replace,'QantasLink','Qantas')) %>% 
  mutate(across(airline,str_replace,"Virgin Australia - Atr/F100 Operations","Virgin Australia")) %>% 
  mutate(across(airline,str_replace,"Virgin Australia Regional Airlines","Virgin Australia"))




```

## Exploratory Data Analysis

### All Major Routes - Total Monthly Pax

Total monthly passenger numbers :

```{r, total_pax}

g <- ind_totals_prep_df %>% 
  #filter(year>2010) %>% 
  ggplot(aes(x=date,y=passenger_trips))+
  geom_line()+
  scale_y_continuous(labels=scales::comma)+
  scale_x_date(date_breaks = "2 year",date_labels = "%y")+
  labs(title = "Australian Domestic Flight History", x="Year", y = "Passenger Numbers (monthly)")+
  theme_bw()
ggplotly(g)

```

Key "points of interest":

-   1987 pilot strike

-   2000 Olympic Games

-   COVID!!!!

Seasonality and trend both also clearly show, at least until covid.

We can break this down by top 10 routes (only tracked 2-way):

### Top 10 Routes - Monthly Pax by O/D City Pairs

```{r,top_routes}

## * Top routes ----
top_routes <- top_routes_prep_df %>% 
  group_by(route,date=max(date)) %>% 
  summarise(passenger_trips = sum(passenger_trips)) %>% 
  ungroup() %>%
  slice_max(passenger_trips, n=10) %>% 
  select(route) %>% 
  pull() %>% 
  as.character()
  

g1 <- top_routes_prep_df %>% 
  filter(route %in% top_routes) %>%
  mutate(route = factor(route,levels=top_routes)) %>% 
  ggplot(aes(date,passenger_trips,colour =route))+
  geom_line()+
  scale_y_continuous(labels=scales::comma)+
  scale_x_date(date_breaks = "2 year",date_labels = "%y")+
  scale_colour_discrete(name  ="Route - 2way")+
  labs(title = "Australian Domestic Flight History - Top10 (2-way) Routes", x="by Month", y = "Passenger Numbers (monthly)")+
  #theme_bw() +
  theme(legend.position="bottom")
ggplotly(g1)

```

### All Routes - Total Monthly Pax - Mapped

Following map shows all routes in 2019 (precovid), thickness of line representiing pax volumes for the year (in this case with a moving monthly timeline to show impact of covid - but does not really work that well). As width of line signifies volumes of passenger trips, Sydney-Melbourne route clearly has thickest line!

```{r, leaflet_mapped}

## * Routes Mapped - Leaflet ----

top_routes_short <- top_routes_prep_df %>%
  filter(year>2019) 
  # group_by(year,city1,city2,city1_lng,city1_lat,city2_lng,city2_lat) %>% 
  # summarise(passenger_trips = sum(passenger_trips))

  leaflet() %>% 
    #addProviderTiles(providers$OpenTopoMap) %>% 
    addTiles() %>%
    addProviderTiles(providers$Esri.WorldStreetMap) %>% 
  addFlows(
    top_routes_short$city1_lng, 
    top_routes_short$city1_lat, 
    top_routes_short$city2_lng, 
    top_routes_short$city2_lat,
    flow = top_routes_short$passenger_trips,
    time = top_routes_short$date,
    dir = 0,
    minThickness = .1,
    maxThickness = 5,
    popupOptions = list(closeOnClick = FALSE, autoClose = FALSE)
  )

```
